Verification of Stencil Computations through Spatial Specifications

### 2016-07-06 paper.tex
edits to section 3

\subsection{Equational theory and approximations}
\label{sec:eqs}

@@

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\label{-0.5em}
\end{figure}

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations for $<:$ on \textit{spec}), and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) axiom connects
equations with inequations. The (\trule{\textsc{over}}) explains
the notion of spatial over approximation via overlapping regions.
The (\trule{\textsc{rule}}) allows the modifier that
enforces that every index is read only once to be dropped
as an approxiamtion. The (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

\anote{Can we say something here about why these equations are useful i.e. we use them to do simplification of stencils?}

@@

We define a semantic model for our stencil specification language.
This model has a number of purposes: (1) it is used in the inference,
checking, and program synthesis algorithms (introduced in
Section~\ref{sec:analysis}); (2) gives a basis for the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model is a denotational
semantics, in terms of sets of vectors which we call \emph{index
  schemes}. We introduce these first and then define the denotational semantics.

@@

\begin{definition}[Index scheme]
An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.
Values in $\mathbb{Z}$ represent offsets
of a neighbourhood index (indexing
expressions of the form $i + a$ where $i$ is an induction variable
and $a$ is a constant offset, Definition~\ref{def:neighbour},
p.~\pageref{def:neighbour}). The additional value
$\infty$ represents any indexing behaviour--- it is a
wildcard.
Throughout $u, v, w$ range over index schemes, and $u_i$ denotes
the $i^{th}$ element of $u$.
\end{definition}

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation forms the basis of the analysis required for the inference
and the checking procedures in Section~\ref{sec:analysis}.

\begin{definition}[Schematic]
The partial function $\textsf{schematic}$ maps
source-language indexing terms $(e_1, \ldots, e_n)$
into $n$-dimensional index schemes
$\in (\domainVal)^n$, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality explicitly in our types with the set $A + \bot$.} function
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
which maps each individal indexing expression to a component of
the indexing scheme, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{} is not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example $\textsf{schematic}(i, j+1) = \vtwoh{0}{1}$ assuming
$i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{definition}

@@

\subsection{Denotational model}

@@

%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form. The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
                                              \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
      \textsf{lower} \, (S \hat{\odot} T) \\
\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
      \textsf{upper} \, (S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u)
\end{align*}
For brevity we omit the cases where the parameters above are flipped
as $\odot$ is commutative.
\end{definition}

>>>>

\subsection{Equational theory and sub-specifications}
\label{sec:eqs}

@@

The equational theory provides not only an axiomatic semantics
for the language, which the user can use to rewrite their
specification, but it is also used extensively in the inference
procedure (Section~\ref{subsec:inference}) to simplify specifications.

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

Figure~\ref{fig:inequations} defines a notion of sub-specifications
via the relation $<:^r$ on \textit{region}
syntax and then $<:$ on \textit{spec} syntax inductively.
Both relations are reflexive and transitive (we omit
these equations) and congruent
with respect to \term{+}, \term{*}, \term{atLeast}, \term{atMost},
and \term{readOnce}. The (\trule{\textsc{eq}}) rule connects region
equations with inequations. The (\trule{\textsc{over}}) rule explains
the notion of spatial over-approximation via overlapping regions.
The (\trule{\textsc{rep}}) allows the modifier that
enforces that every index is read just once to be dropped
as an approxiamtion; (\trule{\textsc{bound}}) gives
a connection between the \term{atLeast} and \term{atMost}
approximation modifiers.

@@

We define a denotational model of the semantics of our 
stencil specification language. This model has a number of purposes: (1) it is used in the inference,
checking, and synthesis algorithms (Section~\ref{sec:analysis}); (2)
justifies the equational
theory and approximations of the previous section; and (3) can be used
to guide correct implementations. The model domain is over sets of vectors which we call \emph{index
  schemes}, which we introduce first before the model in Section~\ref{subsec:model}.

@@

\begin{definition}
An \emph{index scheme} is a vector of size $n$ (an $n$-vector)
with values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.
Values in $\mathbb{Z}$ represent offsets
of a neighbourhood index (indexing
expressions of the form $i + a$ where $i$ is an induction variable
and $a$ is a constant offset, Def.\ref{def:neighbour},
p.\pageref{def:neighbour}). The additional value
$\infty$ represents any indexing behaviour--- it is a
wildcard.
Throughout $u, v, w$ range over index schemes and $u_i$ denotes
the $i^{th}$ element of $u$. We write $(\domainVal)^n$ for
the set of $n$-dimensional schemes.
\end{definition}

Indexing schemes give an abstract
representation of index expressions in the source language.
Our model represents the meaning of stencil specifications
as indexing schemes, which explain the range of possible
indexing behaviours allowed under the specification.
The relationship between indexing schemes and source language
terms is given precisely by the partial function, \textsf{schematic}, which
maps syntactic array-index terms to indexing schemes. This
operation is used by the inference
and the checking procedures.

\begin{definition}%[Schematic]
The partial function $\textsf{schematic}$ maps
source indexing terms $(e_1, .., e_n)$
to $n$-dimensional index schemes, defined:
%
\begin{align*}
\mathsf{schematic}(e_1, \ldots, e_n) & =
[\downarrow\!e_1 \ldots \downarrow\!e_n ]
\quad \textit{iff} \;\; \forall i . \downarrow\!e_i \neq \bot
\end{align*}
%
The $\downarrow$ operator is a partial\footnote{We represent
  partiality in our signatures with the set $A + \bot$.} 
function
 $\downarrow : \textit{expr} \rightarrow (\domainVal + \bot)$
mapping individal indexing expressions to 
an indexing scheme component, defined:
%%
\begin{align*}
\downarrow\!e
 & =  \begin{cases}
a & e \equiv i + a \; \wedge \; \mathsf{neigh}(i + a) \\
-a & e \equiv i - a \; \wedge \; \mathsf{neigh}(i - a) \\
\infty & \textit{$\mathsf{IV}(e) = \emptyset$} \\
\bot   & \textit{otherwise}
\end{cases}
\end{align*}
%%
where $\mathsf{IV}(e)$ is the set of free induction variables
appearing in an expression, and $\neigh{}$ classifies expressions which
are constant offsets from an induction variable
(Defn.~\ref{def:neighbour}). In the third case,
if the expression is not a neighbourhood index but is not
a relative index either (\ie{}, not defined in terms
of an induction variable) then it is mapped to $\infty$, otherwise
$\downarrow\!e$ is undefined. The $\mathsf{schematic}$ function
is undefined if $\downarrow$ of any of its components $e_i$ is
undefined.

For example, assuming $i$ and $j$ are induction variables and $\textsf{schematic}(i-1, 2, j)
 = \vthreeh{-1}{\infty}{0}$. However, $\textsf{schematic}(i*2, j)$, for
 example, is
 undefined as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{definition}

@@

\subsection{Denotational model}
\label{subsec:model}

@@

%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
  is defined in terms of the intermediate $\hat{\odot}$ on \specDNF{}:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\odot$ is essentially applying
 the \trule{\textsc{dist}} rule of our equational theory
 (Figure~\ref{fig:equations}). The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{exact}(S) \odot \textsf{exact}(T) = \; & \textsf{exact}(S
                                              \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{lower}(T) = \; & 
      \textsf{lower} \, (S \hat{\odot} T) \\
\textsf{upper}(S) \odot \textsf{upper}(T) = \; &
      \textsf{upper} \, (S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
      \; \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u)
\end{align*}
For brevity we omit the cases where the parameters above are flipped
as $\odot$ is commutative.
\end{definition}
